# Additional example applications {#additional-examples}

We will present here examples of quantitative bias analysis using simulated participant-level data for a cohort study with a binary treatment, binary outcome, and binary confounder.

```{r, message=FALSE}
# load packages
require(tidyverse)
require(ggplot2)
library(broom) 
library(janitor)
```

```{r, eval=FALSE, include=FALSE}

# define expit function
expit <- function(x) {
  expit_result <- exp(x)/(1 + exp(x))
  return(expit_result)
}

# define sample size
sample_size <- 100000

# define bias parameters
prevalence_confounder_unexp <- 0.25
confounder_treatment_rr <- 2
confounder_outcome_rr <- 2
s11 <- 0.7
s01 <- 1
s10 <- 0.9
s00 <- 1
sensitivity_a0 <-  1
sensitivity_a1 <- 0.8
specificity_a0 <- 1
specificity_a1 <- 0.99

# set seed of random number generator
set.seed(707)
  
# simulate data
sim_data <- tibble(
  # confounder with prevalence 50%
  x = rbinom(sample_size, 1, 0.2), 
  # treatment with prevalence among those without confounder of 25% and with confounder of 50% 
  a = rbinom(sample_size, 1, exp(log(prevalence_confounder_unexp) + log(confounder_treatment_rr)*x)), 
  # outcome 
  y = rbinom(sample_size, 1, expit(log(0.05/0.95) + log(confounder_outcome_rr)*a + log(2)*x)),
  # selection
  s = rbinom(sample_size, 1, s11*a*y + s01*(1-a)*y + s10*a*(1-y) + s00*(1-a)*(1-y)),
  # misclassified outcome
  m_y = rbinom(sample_size, 1, sensitivity_a0*y*(1-a) + sensitivity_a1*y*a + 
                 (1-specificity_a0)*(1-y)*(1-a) + (1-specificity_a1)*a*(1-y))
  )

write_csv(sim_data, "data/simulated_data.csv")
```

## Selection bias

In the simulated data we have a binary treatment $a$, binary confounder $x$, outcome $y$, and a binary indicator for selection into the study $s$. 

```{r, message=FALSE}
sim_data <- read_csv("data/simulated_data.csv") %>% select(x,a,y,s)
sim_data  %>% head(5)
```

The data was generated such that the causal odds ratio between treatment and outcome was 2. However, in a given sample the estimate may differ due to random error. If we observed a random sample from the target population we could unbiasedly estimate the odds ratio using logistic regression.

```{r, message=FALSE}
# fit logistic regression model
lgr_model <- glm(y ~ x + a, data=sim_data, family="binomial") 

# tidy model outputs
or <- lgr_model %>% 
  tidy(exponentiate=TRUE, conf.int=TRUE) %>% 
  filter(term == "a") %>% 
  select(estimate, conf.low, conf.high)

# output as table
or %>% 
  knitr::kable(caption = "Estimated odds ratio")
```
However, if we consider that selection into the study was dependent on exposure and outcome and that we only observed a selected subsample of the target population, then the estimated odds ratio is biased.

```{r}
# restrict data to selected subsample
selected_data <- sim_data %>% filter(s == 1)

# fit logistic regression model
lgr_model_selected <- glm(y ~ x + a, data=selected_data, family="binomial") 

# tidy model outputs
or_selected <- lgr_model_selected %>% 
  tidy(exponentiate=TRUE, conf.int=TRUE) %>% 
  filter(term == "a") %>% 
  select(estimate, conf.low, conf.high) 

# output as table
or_selected %>%
  knitr::kable(caption = "Estimated odds ratio in selected sample")
```

### Bias formulas

One option is to apply bias formulas for the odds ratio.

$$
OR_{BiasAdjusted} = OR_{Observed}\frac{S_{01}S_{10}}{S_{00}S_{11}}
$$
Given that the data was simulated we know the selection probabilities ($S11=0.7$, $S01=1$, $S10=0.9$, $S00=1$) and can directly plug them in to estimate a bias-adjusted odds ratio. However, in practice we will not know these probabilities and will typically specify a range of values, or for probabilistic bias analysis a distribution of values. 

```{r}
# define bias parameters
S11 <- 0.7
S01 <- 1
S10 <- 0.9
S00 <- 1

# apply bias formula
bias_adjusted_or <- or_selected %>% 
  mutate(across(c(estimate, conf.low, conf.high), ~ .x * (S01*S10)/(S11*S00)))

# output as table
bias_adjusted_or %>% 
  knitr::kable(caption = "Bias-adjusted odds ratio")
```

### Weighting

Alternatively, we can weight the individual records by the inverse probability of selection and use bootstapping to calculate a confidence interval

```{r, warning=FALSE, eval=FALSE}
library(boot)

# Add weights
selected_data_with_weights <- selected_data %>% 
  mutate(prob_select = a*y*S11 + (1-a)*y*S01 + a*(1-y)*S10 + (1-a)*(1-y)*S00) %>%
  mutate(inverse_prob = 1/prob_select) 

# define function to estimate weighted odds ratio (needed for bootstrap function)
calculate_weighted_or <- function(weighted_data, i) {
  weighted_lgr <- glm(y ~ x + a, family="binomial", data=weighted_data[i,], weights=inverse_prob) 
  weighted_or <- coef(weighted_lgr)[["a"]] %>% exp()
  return(weighted_or)
}

# set seed of random number generator to ensure reproducibility
set.seed(747)

# bootstrap calculation of confidence intervals
bootstrap_estimates <- boot(selected_data_with_weights, calculate_weighted_or, R=1000)

# calculate bias-adjusted point estimate using entire selected subsample
point_estimate <- calculate_weighted_or(selected_data_with_weights, 1:nrow(selected_data_with_weights))

# calculate percentile bootstrap confidence interval
conf_int <- quantile(bootstrap_estimates$t, c(0.025, 0.975))

# output bias-adjusted estimate
tibble(estimate=point_estimate, conf.low=conf_int[[1]], conf.high=conf_int[[2]]) %>%
  knitr::kable()
```

If we expect selection probabilities to differ within levels of covariates then we can specify different selection probabilities for different strata of covariates.

## Misclassification

We will now consider some quantitative bias analysis methods for misclassification of a binary outcome. Similar approaches can be applied for misclassification of a binary exposure. 

With differential misclassification of the outcome, $m_y$, the odds ratio is biased.

```{r, message=FALSE}
# load data
misclassified_data <- read_csv("data/simulated_data.csv") %>% select(x,a,m_y,s)

# fit logistic regression model with misclassified outcome
lgr_model_misclassified <- glm(m_y ~ x + a, data=misclassified_data, family="binomial") 

# tidy model outputs
or_misclassified <- lgr_model_misclassified %>% 
  tidy(exponentiate=TRUE, conf.int=TRUE) %>% 
  filter(term == "a") %>% 
  select(estimate, conf.low, conf.high) 

# output as table
or_misclassified %>%
  knitr::kable(caption = "Estimated odds ratio with misclassified outcome")
```

### Bias formulas

Bias formulas for misclassification typically apply to 2x2 tables or 2x2 tables stratified by covariates and require us to specify the bias parameters of sensitivity and specificity. Given that the data was simulated, we know that sensitivity and specificity among the treated were 80% and 99%, and that sensitivity and specificity among the unexposed were 100%. In practice, we do not know these values, but can estimate them using validation studies and specify a range or, for probabilistic bias analysis, distribution  of plausible values. 

```{r, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
library(kableExtra)
crude_table <- tibble(lbl=c("Y*=1", "Y*=0"), `A=1`=c("a", "c"), `A=0`=c("b", "d"))
crude_table %>% knitr::kable(col.names=c("", "A=1", "A=0"),
                             caption="Observed 2x2 table") %>% 
  kableExtra::column_spec(1, bold = TRUE)
```

```{r, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
library(kableExtra)
corrected_table <- tibble(lbl=c("Y=1", "Y=0"), `A=1`=c("A", "C"), `A=0`=c("B", "D"))
corrected_table %>% knitr::kable(col.names=c("", "A=1", "A=0"),
                             caption="Corrected 2x2 table") %>% 
  kableExtra::column_spec(1, bold = TRUE)
```

```{r}

# specify bias parameters
sensitivity_a0 <-  1
sensitivity_a1 <- 0.8
specificity_a0 <- 1
specificity_a1 <- 0.99

# define function to correct 2x2 table
correct_two_by_two <- function(a, b, c, d, sensitivity_a0, sensitivity_a1, specificity_a0, specificity_a1) {
  A <- (a - (a + c)*(1-specificity_a1))/(sensitivity_a1 + specificity_a1 - 1)
  B <- (b - (b + d)*(1-specificity_a0))/(sensitivity_a0 + specificity_a0 - 1)
  C <- a + c - A
  D <- b + d - B
  return(c("A"=A,"B"=B,"C"=C,"D"=D))
}

# extract 2x2 table values
a_x0 <- misclassified_data %>% filter(m_y==1, a==1, x==0) %>% nrow()
b_x0 <- misclassified_data %>% filter(m_y==1, a==0, x==0) %>% nrow()
c_x0 <- misclassified_data %>% filter(m_y==0, a==1, x==0) %>% nrow()
d_x0 <- misclassified_data %>% filter(m_y==0, a==0, x==0) %>% nrow()
a_x1 <- misclassified_data %>% filter(m_y==1, a==1, x==1) %>% nrow()
b_x1 <- misclassified_data %>% filter(m_y==1, a==0, x==1) %>% nrow()
c_x1 <- misclassified_data %>% filter(m_y==0, a==1, x==1) %>% nrow()
d_x1 <- misclassified_data %>% filter(m_y==0, a==0, x==1) %>% nrow()

# correct 2x2 table 
corr_two_by_two_x0 <- correct_two_by_two(a_x0, b_x0, c_x0, d_x0, sensitivity_a0, sensitivity_a1, specificity_a0, specificity_a1)
corr_two_by_two_x1 <- correct_two_by_two(a_x1, b_x1, c_x1, d_x1, sensitivity_a0, sensitivity_a1, specificity_a0, specificity_a1)

# extract corrected 2x2 table values
A_x0 <- corr_two_by_two_x0[[1]]
B_x0 <- corr_two_by_two_x0[[2]]
C_x0 <- corr_two_by_two_x0[[3]]
D_x0 <- corr_two_by_two_x0[[4]]
A_x1 <- corr_two_by_two_x1[[1]]
B_x1 <- corr_two_by_two_x1[[2]]
C_x1 <- corr_two_by_two_x1[[3]]
D_x1 <- corr_two_by_two_x1[[4]]

# calculate Mantel-Haenszel odds ratio
numerator <- A_x1*D_x1/(A_x1 + B_x1 + C_x1 + D_x1) + A_x0*D_x0/(A_x0 + B_x0 + C_x0 + D_x0)
denominator <- B_x1*C_x1/(A_x1 + B_x1 + C_x1 + D_x1) + B_x0*C_x0/(A_x0 + B_x0 + C_x0 + D_x0)

mh_estimate <- numerator/denominator
print(mh_estimate)
```

We can use bootstrapping to calculate a confidence interval.

### Record-level correction

For record-level correction we can as before calculate two by two tables, but as a second step use these to calculate predictive values which we can use to impute a *corrected* exposure.

```{r}
# calculate predictive values
ppv_a1_x1 <- sensitivity_a1*A_x1/a_x1
ppv_a0_x1 <- sensitivity_a0*B_x1/b_x1
npv_a1_x1 <- specificity_a1*C_x1/c_x1
npv_a0_x1 <- specificity_a0*D_x1/d_x1 
ppv_a1_x0 <- sensitivity_a1*A_x0/a_x0
ppv_a0_x0 <- sensitivity_a0*B_x0/b_x0
npv_a1_x0 <- specificity_a1*C_x0/c_x0
npv_a0_x0 <- specificity_a0*D_x0/d_x0 

# impute outcome
misclassified_data <- misclassified_data %>% mutate(y = rbinom(100000, 1, ppv_a1_x1*a*m_y*x + ppv_a0_x1*(1-a)*m_y*x + (1-npv_a1_x1)*a*(1-m_y)*x + (1-npv_a0_x1)*(1-a)*(1-m_y)*x + ppv_a1_x0*a*m_y*(1-x) + ppv_a0_x0*(1-a)*m_y*(1-x) + (1-npv_a1_x0)*a*(1-m_y)*(1-x) + (1-npv_a0_x0)*(1-a)*(1-m_y)*(1-x)))

# fit a logistic regression model using imputed exposure
glm(y ~ a + x, family="binomial", data=misclassified_data) %>% 
  tidy(exponentiate=TRUE) %>% filter(term == "a") %>% select(estimate)
```

### Probabalistic bias analysis



## Unmeasured confounding


